{\scshape Denicek}: Computational Substrate for\\ Document-Oriented End-User Programming

### 2024-12-08 paper.tex
Experiences, methodology

% TODO: Figure out how to call those "programming experiences" (also is this a good name?)
%   the text uses "concrete, collaborative, interactive" - but this is a silly list
%   really, they are "cool" but how to say that.
%
% TODO: Also, what exactly goes into the list?
%
%
\documentclass[sigconf]{acmart}
%\documentclass[sigconf,review,anonymous]{acmart}

@@

\setcopyright{acmlicensed}
\copyrightyear{2018}
\acmYear{2018}
\acmDOI{XXXXXXX.XXXXXXX}

\acmConference[Conference acronym 'XX]{Make sure to enter the correct
  conference title from your rights confirmation emai}{June 03--05,
  2018}{Woodstock, NY}
\acmISBN{978-1-4503-XXXX-X/18/06}

\begin{document}
\title[Denicek: Computational Substrate for Concrete, Collaborative, Interactive
  Programming]{{\scshape Denicek}: Computational Substrate for Concrete, \\Collaborative, Interactive Programming}

@@

\begin{abstract}
Research on interactive programming systems gave rise to a range of compelling programming
experiences, including local-first collaborative editing, programming by demonstration,
incremental recomputation, schema and code co-evolution, provenance tracking and managed copy and
paste. Those experiences advance the state of the art of programming, but they are hard to
implement on the basis of existing programming languages and systems.

We contribute Denicek, a computational substrate that reduces the complexity of
implementing the aforementioned programming experiences. Denicek represents a program as a series of edits
that construct and transform a document consisting of data and formulas. Denicek provides two
primitive operations on series of edits, merging and conflict checking, that form the backbone
of the implementation of a range of programming experiences.

We discuss the architecture of Denicek, document key design considerations and elaborate
the implementation of the programming experiences listed above. To evaluate the proposed
architecture, we use Denicek as the basis of a simple innovative data exploration environment.
The case study shows that the Denicek computational substrate provides an appealign basis for the
design of richer and more accessible interactive programming systems.
\end{abstract}

\keywords{Programming Systems, Computational Substrate, Programming by Demonstration,
  Local-First Software}

@@

@@

A computational substrate defines the basic structures using which programs are constructed, how
the progam state is represented and how it is changed \cite{jakubovic-2022-ladder}. The choice of
a substrate affects what programming experiences can be readily supported. For example,
object-oriented programming has been historically linked to graphical user
interfaces~\cite{kay-1993-smalltalk}, while using lists as the basic substrate enabled Lisp to become
a ``language laboratory'' \cite{steele-1993-lisp}.

In principle, any programming experience can be developed on top of any computational
substrate. However, a program representation and operations for working with programs provided by
a suitable substrate can eliminate much of the complexity of implementing an interesting
programming experience. For example, the reflective capabilities of Smalltalk make it easy
to provide rich debugging tools \cite{rauch-2019-babylonian} that would be difficult to provide
for C/C++~\cite{kell-2024-debugging}.

In this paper, we ask what would be the ideal programming substrate
for making programming easier, collaborative and more transparent? We seek a programming
substrate that makes it easy to develop programming experiences such as:

\begin{itemize}
\item \emph{Programming by demonstration} --
  Allow non-programmers to construct simple programs by performing examples of the expected behaviour. \cite{leiva-2021-rapido,cypher-1993-pbd}.
\item \emph{Local-first collaboration} --
  Multiple users should be able to use and modify a single program, preferrably without requiring a central server. \cite{kleppmann-2019-local}
\item \emph{Provenance tracking} --
  The execution of the program should leave an understandable trace that lets the user understand why program resulted in a particular result.
\item \emph{Schema evolution [extra-ish]} --
  When the user evolves the structure of the program, data and code should co-evolve automatically to match the new structure.
\item \emph{Notational freedom [extra-ish?]} --
  Allow users to adapt the program using a notation that suits them and is appropriate for the programming task at hand. [Joel]
\item \emph{Concrete programming [extra?]} --
  It should be possible to reuse parts of program or program logic without constructing abstractions, for example by managed copy \& paste.\cite{edwards-2006-copypaste,edwards-2022-copypaste}
\end{itemize}

\subsubsection*{Methodology}

DESIGN PROCESS = formative examples + evaluation case study

@@

@@

\subsection{Programming Experiences}
list

\subsection{Substrate}
how it works very roughly

\subsection{Contributions}
one main thing - substrate - with other things

@@

@@

>>>>

% TODO: Figure out how to call those "programming experiences" (also is this a good name?)
%   the text uses "concrete, collaborative, interactive" - but this is a silly list
%   really, they are "cool" but how to say that.
%
% TODO: Also, what exactly goes into the list?
%
%
\documentclass[sigconf,anonymous]{acmart}
%\documentclass[sigconf,review,anonymous]{acmart}

@@

\setcopyright{none}
\copyrightyear{2025}
\acmYear{2025}
\acmDOI{XXXX.XXXX}

\acmConference[Conference acronym 'XX]{}{June 03--05, 2018}{Woodstock, NY}
\acmISBN{978-1-4503-XXXX-X/18/06}

\begin{document}
\title[Denicek: Computational Substrate for Document-Oriented End-User
  Programming]{{\scshape Denicek}: Computational Substrate for\\ Document-Oriented End-User Programming}

@@

\begin{abstract}
Human-computer interaction research gave rise to a wide range of compelling programming experiences,
including local-first collaborative editing, programming by demonstration, incremental
recomputation, schema change control, end-user debugging and concrete programming.
Those experiences advance the state of the art of end-user programming, but they are hard to
implement on the basis of established programming languages and systems.

We contribute Denicek, a computational substrate that reduces the complexity of
implementing the aforementioned programming experiences. Denicek represents a program as a series of edits
that construct and transform a document consisting of data and formulas. Denicek provides three
primitive operations on series of edits, application, merging of histories and conflict checking.
We show that the three operations form the backbone of an easy implementation of a range of
programming experiences.

We discuss the architecture of Denicek, document key design considerations and elaborate
the implementation of the programming experiences listed above. To evaluate the proposed
architecture, we use Denicek as the basis of a simple innovative data exploration environment.
The case study shows that the Denicek computational substrate provides an appealing basis for the
design of rich, interactive end-user programming systems.
\end{abstract}

\keywords{Programming Systems, Computational Substrate, End-User Programming,
  Programming by Demonstration, Local-First Software}

@@

A computational substrate defines the basic structures using which programs are constructed, how
the progam state is represented and how it is changed \cite{jakubovic-2022-ladder}. The choice of
a substrate affects what programming experiences can be readily supported by a system. For example,
object-oriented programming has been historically linked to graphical user
interfaces~\cite{kay-1993-smalltalk}, while using lists as the basic substrate enabled Lisp to become
a language laboratory \cite{steele-1993-lisp}.

In principle, any programming experience can be developed on top of any computational
substrate. However, a suitable program representation and operations for working with programs
provided by the substrate can eliminate much of the complexity of implementing interesting
programming experiences. For example, the reflective capabilities of Smalltalk make it easy
to provide rich debugging tools \cite{rauch-2019-babylonian} that are difficult to provide
for C/C++~\cite{kell-2024-debugging}.

What would be the ideal programming substrate for rich, interactive end-user programming systems?
In this paper, we aim to design a programming substrate that makes it easy to develop a range of
programming experiences \cite{myers-2006-eup}:

\begin{itemize}
\item \emph{Programming by Demonstration.} Allow users to construct simple programs by
  demonstrating the steps of the expected behavior using concrete examples~\cite{leiva-2021-rapido,cypher-1993-pbd}.
\item \emph{Local-First Collaboration.} Multiple users should be able to concurrently
  modify a single document and merge their changes, preferrably without requiring a central server~\cite{kleppmann-2019-local,litt-2022-peritext}.
\item \emph{End-User Debugging.} The user should be able to ask provenance questions~\cite{cheney-2009-provenance} --
  why a computation resulted in a particular value and what inputs contributed to the result \cite{ko-2009-whyline}.
\item \emph{Schema Change Control.} When the user evolves the structure of the document, affected
  data and formulas should automatically co-evolve to match the new structure~\cite{litt-2020-cambria,edwards-2025-schema}.
\item \emph{Concrete Programming.} It should be possible to reuse parts of program logic, or formulas,
  without introducing abstractions, that is, program against concrete values \cite{edwards-2006-copypaste,edwards-2022-copypaste}.
\end{itemize}

\subsubsection*{Methodology}
We use a two-stage research methodology. In the first stage, we identify a number
of \emph{formative examples} -- simple programming tasks, drawn from literature, that manifest one or
more of the desired programming experiences. Using those examples, we co-design the Denicek substrate
and Webnicek (Section~\ref{sec:walk}), a simple web-based end-user programming environment based on the substrate.
Although Webnicek can be used to complete end-user programming tasks, it is optimized for
testing the underlying substrate rather than for usability.

In the second stage, we use the Denicek substrate as the basis of Datnicek, an end-user data
exploration environment inspired by existing research systems\cite{kandel-2011-wrangler,drossos-2020-wrex}.
We use the second stage to evaluate suitability of the Denicek substrate for the development of
end-user programming systems and report the results of our evaluation (Section~\ref{sec:eval}).

\subsubsection*{Substrate}
how it works very roughly

\subsubsection*{Contributions}
one main thing - substrate - with other things

@@

% The problems include a counter app \cite{kiss-2014-7guis},
% Todo list app and conference planning \cite{edwards-2025-schema}.

@@

% debugging
% https://dl.acm.org/doi/pdf/10.1145/1133265.1133293

@@

METHODOLOGY (add somewhere belo) - table of "formative examples" and experiences they require?
+ SOURCE (literature?)

@@

what is this about?
end-user programming ? Cite % https://dl.acm.org/doi/pdf/10.1145/1125451.1125472
but also Live, Interactive etc. - perhaps for a lack of better term use end-user
(also notebooks)

@@

\section{Evaluation}
\label{sec:eval}
how well did Denicek work for implementing Datnicek
